---
title: "Analysis of NZTA crash data"
author: "Jayden Mudge"
date: "07/05/2024"
output: slidy_presentation
---

```{r setup, include=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)
library(randomForest)
library(glmnet)
library(VIM)
library(fastDummies)
opts_chunk$set(echo = FALSE)
opts_knit$set(root.dir = paste0(getwd(), "/.."))
```

## Overview

What I don't aim to do:

- Fully identify and resolve the root causes of all crashes in Aotearoa New Zealand

## Overview

What I don't aim to do:

- Fully identify and resolve the root causes of all crashes in Aotearoa New Zealand

Rather, I aim to:

- Showcase my analytical skills and processes
- Outline some preliminary observations
- Discuss other ideas to explore

## Initial explorations

Let's start off by downloading the crash dataset and having a peak at what data is available in it


```{r load_crash_data, echo = FALSE}
crash_csv <- data.table::fread(
  "Data/Crash_Analysis_System_(CAS)_data.csv",
  keepLeadingZeros = TRUE
)


summary(crash_csv[,1:8])
```


## What do we find?

There's a lot of fascinating data in here!

- X,Y co-ordinates, street names, and other geographic info
- Both calendar and financial year, and holidays if applicable
- Vehicle counts involved in crash
- Objects crashed into, where applicable
- Weather and light conditions
- Outcomes on individuals: injuries and severity

## An initial visualisation

```{r reg_vis, echo = FALSE, out.width = '100%', fig.height = 5}
vehicle_vars <- c(
  "bicycle", "bus", "carStationWagon", "moped", "motorcycle", "otherVehicleType", "schoolBus",
  "suv", "taxi", "train", "truck", "unknownVehicleType", "vanOrUtility", "vehicle"
)

obstacle_vars <- c(
  "bridge", "cliffBank", "debris", "ditch", "fence", "guardRail", "houseOrBuilding",
  "kerb", "NumberOfLanes", "objectThrownOrDropped", "otherObject", "overBank",
  "parkedVehicle", "pedestrian", "phoneBoxEtc", "postOrPole", "roadworks", "slipOrFlood",
  "strayAnimal", "trafficIsland", "trafficSign", "tree", "waterRiver"
)

outcome_vars <- c("fatalCount", "seriousInjuryCount", "minorInjuryCount")

na_to_0 <- function(x) {
  coalesce(x, 0)
}

#impute missing speed limits
crash_csv <- VIM::matchImpute(
  crash_csv,
  variable = "speedLimit",
  match_var = c("crashLocation1", "crashLocation2", "NumberOfLanes", "crashSHDescription")
  )

crash_csv <- crash_csv %>%
  mutate(
    across(all_of(c(vehicle_vars, obstacle_vars, outcome_vars)), na_to_0),
    advisorySpeed = coalesce(advisorySpeed, speedLimit),
    tempChangeInSpeedLimit = ifelse(is.na(temporarySpeedLimit), 0, speedLimit - temporarySpeedLimit),
    tlaId = stringr::str_pad(tlaId, width = 3, pad = "0"),
    involved_motorcycle = (motorcycle > 0),
    severeCrash = (crashSeverity %in% c("Serious Crash", "Fatal Crash"))
  )

region_order <- crash_csv %>% 
  arrange(tlaId) %>% 
  distinct(region) %>% 
  pull()

region_order <- c(
  "Northland Region", "Auckland Region", "Waikato Region", "Bay of Plenty Region", "Gisborne Region",
  "Hawke's Bay Region", "Taranaki Region", "Manawatū-Whanganui Region", "Wellington Region", "Tasman Region",          
  "Nelson Region", "Marlborough Region", "West Coast Region", "Canterbury Region", "Otago Region", "Southland Region", ""
  )

ggplot(crash_csv %>% filter(region != "")) + 
  geom_histogram(aes(x=crashYear, fill = crashSeverity), binwidth = 1) +
  facet_wrap(~region, axes = "all_x") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Observations

There are some limitations with this data source though

- No info on traffic volumes to provide context to crash numbers. Should one avoid Auckland because it's where the majority of crashes happen?
- No date-time information more precise than year
- Which parties were responsible for the crash, and which parties received the injuries

We also seem to have incomplete data for 2023, perhaps we should drop this as it skews our analysis.

Also notice a dip in crashes between 2010 and 2015 in all regions, what could be driving that? A [report from NZTA](https://www.transport.govt.nz/assets/Uploads/Presentation/Overview-of-Road-Safety-in-NZ-Data-packs-for-reference-groups.pdf) noted the increase in fatalities could be driven by an increase in motorcycle registrations.

## Crashes involving motorbikes

```{r severe_pct_reg, echo = FALSE, out.width = '100%'}
ggplot(crash_csv %>% filter(crashYear < 2023)) + 
  geom_bar(aes(x=crashYear, fill = involved_motorcycle)) +
  facet_wrap(~crashSeverity, scales = "free_y")
```

However, this doesn't seem to be a major contributor to overall traffic collisions. The main drivers in fact appear to be minor and non-injury crashes. Perhaps it's just an increase in rates of these minor incidents being reported?

## Applying context

Using the high quality population estimates from Stats NZ, we can calculate per 1,000 usual residents for crashes at each severity level. I also include a per 1,000 people of driving age measure, as these are the people most likely to be behind the wheel of a car.

```{r cars, echo = FALSE}
erp_by_ta <- data.table::fread(
  "Data/erp_by_ta.csv", skip = 1
)

erp_by_ta <- erp_by_ta %>% 
  fill(
    c('Area', 'Age', 'Sex'), .direction = "down"
  ) %>% 
  filter(
    !is.na(`Year at 30 June`)
  ) %>% 
  mutate(
    tlaId = stringr::str_split_i(Area, pattern = ": ", i = 1),
    pop_group = ifelse(
      Age == "Total people, age", "est_res_pop", "driving_age_pop"
    )
  ) %>% 
  group_by(tlaId, Year = `Year at 30 June`, pop_group) %>% 
  summarise(
    count = sum(Figure)
  ) %>% 
  ungroup() %>% 
  pivot_wider(
    id_cols = c(tlaId, Year),
    names_from = pop_group,
    values_from = count
  )


# Aggregate crash data and combine with population ------------------------

crash_outcomes_by_year <- 
  crash_csv %>% 
  group_by(tlaId, tlaName, crashYear) %>% 
  summarise(
    across(all_of(outcome_vars), sum)
  ) %>%  
  pivot_longer(
    cols = all_of(outcome_vars),
    names_to = "outcome",
    values_to = "n"
  ) %>% 
  mutate(
    outcome = case_when(
      outcome == "minorInjuryCount" ~ "Minor injuries",
      outcome == "seriousInjuryCount" ~ "Serious injuries",
      outcome == "fatalCount" ~ "Fatalities"
    ),
    outcome = factor(outcome, levels = c("Minor injuries", "Serious injuries", "Fatalities"))
  )

crashes_per_capita <- 
  crash_outcomes_by_year %>% 
  inner_join(
    erp_by_ta,
    join_by(tlaId, crashYear == Year),
    relationship = "many-to-many"
  ) %>% 
  mutate(
    outcome_per_cap = n / est_res_pop,
    outcome_per_dap = n / driving_age_pop
  ) %>% 
  filter(crashYear != 2023) #2023 data is not complete so skews picture


# Graphs and tables -------------------------------------------------------

fatalities_per_cap_2022 <-
  crashes_per_capita %>% 
  filter(crashYear == 2022, outcome == "Fatalities") %>% 
  arrange(desc(outcome_per_cap))

serious_injuries_per_cap_2022 <-
  crashes_per_capita %>% 
  filter(crashYear == 2022, outcome == "Serious injuries") %>% 
  arrange(desc(outcome_per_cap))

serious_or_fatal_per_cap <-
  crashes_per_capita %>% 
  filter(outcome %in% c("Fatalities", "Serious injuries")) %>% 
  group_by(tlaId, tlaName) %>% 
  summarise(
    average_per_1000cap = 1000*weighted.mean(x = outcome_per_cap, w = est_res_pop),
    average_per_1000dap = 1000*weighted.mean(x = outcome_per_dap, w = est_res_pop)
  ) %>%
  ungroup() %>% 
  arrange(desc(average_per_1000cap))

head(serious_or_fatal_per_cap) %>% kable()
tail(serious_or_fatal_per_cap) %>% kable()
```

## Statistical modelling for significant variables

```{r ridge_lasso, echo = FALSE}
mod_vars <- c(
  "bicycle", "bridge", "bus", "carStationWagon", "cliffBank", "crashSHDescription",
  "debris", "directionRoleDescription", "ditch", "fence", "flatHill", "guardRail",
  "holiday", "houseOrBuilding", "kerb", "light", "moped", "motorcycle",
  "NumberOfLanes", "objectThrownOrDropped", "otherObject", "otherVehicleType", "overBank",
  "parkedVehicle", "pedestrian", "phoneBoxEtc", "postOrPole", "region", "roadCharacter",
  "roadLane", "roadSurface", "roadworks", "schoolBus", "slipOrFlood", "speedLimit",
  "strayAnimal", "streetLight", "suv", "taxi", "tempChangeInSpeedLimit", "trafficControl",
  "trafficIsland", "trafficSign", "train", "tree", "truck", "urban", "vanOrUtility",
  "waterRiver", "weatherA", "weatherB"
)

character_vars <- crash_csv %>% 
  select(all_of(mod_vars)) %>% 
  select(where(is.character)) %>% 
  names()

mod_data <- crash_csv %>%
  select(all_of(mod_vars)) %>%
  fastDummies::dummy_cols(character_vars, remove_selected_columns = T) %>% 
  as.matrix()

crash_mod <- glmnet::cv.glmnet(
  x = mod_data,
  y = crash_csv$severeCrash,
  family = "binomial",
  alpha = 1,
  intercept = FALSE
)

plot(crash_mod)

coefs <- coef(crash_mod, s = "lambda.1se")

coefs <- data.frame(
  var = coefs@Dimnames[[1]][coefs@i + 1],
  coef = coefs@x
) %>% 
  arrange(desc(coef))
```

```{r key_vars, echo = FALSE}
head(coefs, 10) %>% kable()
tail(coefs, 10) %>% kable()
```